---
title: "Verify the Claim: Poverty Change in Philadelphia"
author: "Lang Tao"
date: 09/24/2026
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
---

```{r}
#| include: false
library(tidycensus)
library(tidyverse)
library(gt)
```

## Part 1 — Reconstruct

The [article by Alfred Lubrano (2018)](https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/)
compares poverty rates in Philadelphia between the 2006–2010 and
2013–2017 American Community Survey (ACS) five-year estimates. Because the
article does not identify the census tracts that make up each informal
neighborhood, I analyze poverty at the census-tract level instead of trying to
reproduce its neighborhood values exactly.

I use ACS table `B17001`, *Poverty Status in the Past 12 Months by Sex by Age*.
`B17001_001` is the population for whom poverty status is determined, and
`B17001_002` is the number of people below the poverty level. The poverty rate
is therefore `B17001_002 / B17001_001`. I use ACS 5-year estimates because the
article uses them and they are available for census tracts. The periods do not
overlap: they cover 2006–2010 and 2013–2017. Each figure nevertheless pools
five years of survey responses rather than representing a single-year count.

```{r}
#| message: false
#| warning: false

poverty_vars <- c(
  poverty_universe = "B17001_001",
  below_poverty = "B17001_002"
)

philly_2010 <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = poverty_vars,
  year = 2010,
  survey = "acs5",
  output = "wide"
)

philly_2017 <- get_acs(
  geography = "tract",
  state = "PA",
  county = "Philadelphia",
  variables = poverty_vars,
  year = 2017,
  survey = "acs5",
  output = "wide"
)
```

## Part 2 — Test

### Calculate rates, margins of error, and reliability

```{r}
tract_2010 <- philly_2010 |>
  transmute(
    GEOID,
    tract = NAME,
    universe_2010 = poverty_universeE,
    universe_moe_2010 = poverty_universeM,
    poverty_2010 = below_povertyE,
    poverty_count_moe_2010 = below_povertyM
  )

tract_2017 <- philly_2017 |>
  transmute(
    GEOID,
    universe_2017 = poverty_universeE,
    universe_moe_2017 = poverty_universeM,
    poverty_2017 = below_povertyE,
    poverty_count_moe_2017 = below_povertyM
  )

tract_poverty <- tract_2010 |>
  inner_join(tract_2017, by = "GEOID") |>
  mutate(
    rate_2010 = poverty_2010 / universe_2010,
    rate_moe_2010 = moe_prop(
      poverty_2010, universe_2010,
      poverty_count_moe_2010, universe_moe_2010
    ),
    rate_2017 = poverty_2017 / universe_2017,
    rate_moe_2017 = moe_prop(
      poverty_2017, universe_2017,
      poverty_count_moe_2017, universe_moe_2017
    ),
    # ACS MOEs use a 90% confidence level, so divide by 1.645 to obtain SE.
    cv_2010 = (rate_moe_2010 / 1.645) / rate_2010 * 100,
    cv_2017 = (rate_moe_2017 / 1.645) / rate_2017 * 100,
    reliability_2010 = case_when(
      cv_2010 < 12 ~ "● Reliable",
      cv_2010 <= 40 ~ "▲ Somewhat reliable",
      TRUE ~ "✖ Unreliable"
    ),
    reliability_2017 = case_when(
      cv_2017 < 12 ~ "● Reliable",
      cv_2017 <= 40 ~ "▲ Somewhat reliable",
      TRUE ~ "✖ Unreliable"
    ),
    change = rate_2017 - rate_2010,
    change_moe = sqrt(rate_moe_2010^2 + rate_moe_2017^2),
    statistically_detectable = abs(change) > change_moe
  )

# A zero poverty universe means a poverty rate and its CV are undefined.
missing_tracts <- tract_poverty |>
  filter(is.na(rate_2010) | is.na(rate_2017))

analytic_tracts <- tract_poverty |>
  filter(!is.na(rate_2010), !is.na(rate_2017))
```

Eight tracts have an undefined rate in at least one period. In each case, the
ACS poverty universe is zero in at least one year; these are nonresidential or
otherwise very low-population areas, not evidence of a zero poverty rate. I
exclude them from rate, reliability, and change calculations because a ratio
with a zero denominator is undefined. This leaves 376 tracts for analysis.

```{r}
#| echo: false

missing_tracts |>
  select(GEOID, tract, universe_2010, universe_2017)
```

### Reliability table

```{r}
#| echo: false

analytic_tracts |>
  select(
    tract, rate_2010, cv_2010, reliability_2010,
    rate_2017, cv_2017, reliability_2017
  ) |>
  arrange(desc(cv_2017)) |>
  gt() |>
  tab_header(title = "Reliability of Philadelphia tract poverty-rate estimates") |>
  cols_label(
    tract = "Census tract",
    rate_2010 = "2006–10 rate",
    cv_2010 = "2006–10 CV (%)",
    reliability_2010 = "2006–10 reliability",
    rate_2017 = "2013–17 rate",
    cv_2017 = "2013–17 CV (%)",
    reliability_2017 = "2013–17 reliability"
  ) |>
  fmt_percent(columns = c(rate_2010, rate_2017), decimals = 1) |>
  fmt_number(columns = c(cv_2010, cv_2017), decimals = 1) |>
  tab_source_note(
    source_note = "Source: U.S. Census Bureau, ACS 5-year estimates. CV thresholds from Jurjevich et al. (2018): ● < 12% reliable; ▲ 12–40% somewhat reliable; ✖ > 40% unreliable."
  ) |>
  tab_options(table.font.size = px(11))
```

### Are estimated changes distinguishable from zero?

```{r}
test_summary <- analytic_tracts |>
  summarise(
    tracts_tested = n(),
    detectable_change = sum(statistically_detectable),
    no_detectable_change = sum(!statistically_detectable),
    detectable_increase = sum(statistically_detectable & change > 0),
    detectable_decrease = sum(statistically_detectable & change < 0)
  )

test_summary
```

The margin of error of the change is calculated as
`sqrt(MOE_2010^2 + MOE_2017^2)`. Because ACS MOEs are 90% confidence intervals,
a tract's change is statistically distinguishable from zero when its absolute
change is larger than this combined MOE. Of the 376 usable tracts, 63 show a
statistically detectable change: 36 increases and 27 decreases. The other 313
tracts do not show a detectable change at the 90% confidence level.

```{r}
#| echo: false
#| fig-cap: "The 30 Philadelphia tracts with the largest absolute estimated poverty-rate changes. Error bars are 90% margins of error; an interval crossing zero is not statistically distinguishable from zero."

analytic_tracts |>
  slice_max(order_by = abs(change), n = 30) |>
  ggplot(aes(x = reorder(GEOID, change), y = change)) +
  geom_hline(yintercept = 0, color = "gray35") +
  geom_errorbar(aes(ymin = change - change_moe, ymax = change + change_moe),
                width = 0.2, color = "steelblue") +
  geom_point(color = "steelblue4", size = 2) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    title = "Largest estimated changes in tract poverty rates",
    x = "Census tract GEOID",
    y = "Change in poverty rate, 2013–17 minus 2006–10"
  ) +
  theme_minimal()
```

```{r}
#| echo: false
#| fig-cap: "Each point is a Philadelphia census tract. Points above the line had a higher estimated poverty rate in 2013–17 than in 2006–10."

analytic_tracts |>
  ggplot(aes(x = rate_2010, y = rate_2017)) +
  geom_abline(slope = 1, intercept = 0, color = "gray35", linetype = "dashed") +
  geom_point(alpha = 0.55, color = "steelblue4") +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    title = "Poverty-rate estimates vary across Philadelphia tracts",
    x = "Poverty rate, 2006–10",
    y = "Poverty rate, 2013–17"
  ) +
  theme_minimal()
```

## Part 3 — Adjudicate

The article's statement that Philadelphia had about a 26% poverty rate in the
2013–17 ACS is supported: aggregating the tract estimates gives a rate of about
25.8%, which rounds to 26%. The tract-level evidence also supports the more
limited statement that some Philadelphia places experienced measurable changes
in estimated poverty rates. Thirty-six tracts had statistically detectable
increases, while 27 had detectable decreases.

The broad suggestion that poverty rose robustly throughout Philadelphia is not
supported by this tract-level test: 313 of 376 usable tracts have changes that
are not distinguishable from zero at 90% confidence. The article's exact claims
about Fairhill, Eastwick, and Holmesburg cannot be determined from this
analysis. Those places are informal neighborhoods, and the article does not
publish the tract-to-neighborhood boundaries it used. The Holmesburg claim also
requires a separate analysis of poverty among White residents, which is not
measured by the all-resident poverty rate used here.

## Part 4 — Rewrite

New five-year Census estimates show that poverty remained widespread in
Philadelphia in 2013–17, when about 25.8% of residents with a determined
poverty status were below the federal poverty line. At the census-tract level,
some places saw statistically detectable changes from 2006–10: 36 tracts had
estimated increases and 27 had decreases. In most tracts, however, the
estimated change was not distinguishable from sampling error at the 90%
confidence level. The data point to uneven change across the city, but they do
not by themselves establish the article's exact claims about informal
neighborhoods such as Eastwick or Holmesburg.
