---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Lang Tao"
date: 2026/09/08
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---
## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question
Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

**In this analysis, I look at median gross rent across counties in New York, and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure

```{r}
my_state <- "NY"        # two-letter abbreviation, in quotes
my_variable <- "B25064_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**The dataset contains 62 rows, matching the 62 counties in New York State.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```
**[Hamilton County has the largest margin of error relative to its estimated median gross rent. Its estimate is 860 dollars with a margin of error of 95 dollars, resulting in a relative margin of error of approximately 11.05%.]**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

**[Using the original reliability thresholds, 51 counties are classified as high confidence, 10 as moderate, and one as low confidence]**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```


**[High-confidence counties have an average population of approximately 381,479, while moderate-confidence counties have an average population of approximately 37,898. No counties fall into the low-confidence category under these thresholds. This pattern suggests that more populous counties tend to have more reliable median gross rent estimates than less populous counties.]**

---

## What this means for policy

**1.Among New York counties, we know the least about housing costs in Hamilton County, which has the largest relative margins of error at approximately 11.05%. In addition, Schuyler County, Greene County and Lewis County have high margin of error, at 8.74%, 7.96% and 7.19%, which showes that madian gross rent is not accurate for these countries.
   2. This pattern occurs because the ACS is based on a sample rather than a complete count of every resident. Smaller counties get fewer responses. So according to our math knowledge, Its results are less accurate.
   3. If we assume that every piece of data is highly accurate, then these very small counties might receive too much or too little funding, leading to either overspending or underspending—which could result in wasted funds or people not receiving sufficient assistance. Rural counties with very small populations, such as Hamilton County, are particularly vulnerable.**

---

